Cluster Difference Imaging Photometric Survey. III. An Open Cluster Spread Across 500\,pc

### 2021-03-13 cluster_paper/ms.tex
update ms
with lithium

\begin{figure}[t]
	\begin{center}
		\leavevmode
		%\subfloat{
			\includegraphics[width=0.49\textwidth]{f5a.pdf}
		%	\includegraphics[width=0.59\textwidth]{f5b.pdf}
		%}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown
  versus extinction-corrected color for all kinematic \cn\ members
  with Gaia-ESO ($N=452$) or GALAH ($N=107$) spectra available.
  The GALAH spectra comprise slightly over half of the halo stars,
  due to the non-targeted selection function of that survey.
  Points with ${\rm EW}\approx0\,$m\AA\ are non-detections.
  \label{fig:lithiumcorehalo}
  }
\end{figure}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		%\subfloat{
		\includegraphics[width=1\textwidth]{f5b.pdf}
		%	\includegraphics[width=0.59\textwidth]{f5b.pdf}
		%}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Lithium and rotation in NGC\,2516.}
		Points show all kinematic \cn\ members with TESS rotation
		periods in the Gaia-ESO or GALAH surveys, and are colored by
		equivalent width.
	   Non-detections appear with ${\rm EW}\approx0\,$m\AA.
	The K-dwarfs shows a similar lithium-rotation correlation
		to what has been previously observed,
		and is discussed in Section~\ref{subsec:lithium}.
		\label{fig:lithium}
	}
\end{figure*}

@@

Figure~\ref{fig:lithium} shows the results.  
At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.
The fact that we see the same trend across the core and halo of \cn\ 
{\it a)} supports the conclusion that the halo is coeval with the
core, and {\it b)} suggests that the lithium-rotation correlation is
not explained by environmental effects such as the density of the cluster
during its embedded phase, but is instead tied to physical processes
acting on all low-mass pre-main-sequence stars.

% TODO: this paragraph more or less parrots the processes described by
% Bouvier+20. are there others? is there a better way to describe
% these?
Processes both internal and external to the star have been suggested to explain
the lithium-rotation trend (CITE see the recent review by Bouvier+20).
One explanation based in the stellar interior could be that the
convective mixing efficiency is anticorrelated with the surface
rotation (e.g., CITE Siess + Livio 1997, Baraffe+2017).  Another
possibility could be that stronger magnetic fields in the star's
interior inhibit convection (READ e.g., Ventura+98, Chabrier+07,
Somers + Pinsonneault 2014).  An external process that might also be
important is the effect of star-disk magnetic locking during the PMS
phase (CITE: magnetic braking).  Longer disk lifetimes would lead to
the star's outer convective zone being locked for longer while the
radiative core contracts.  The resulting differential rotation and
rotational mixing could drive the lithium depletion (CITE: Bouvier 08,
Eggenberger+12).
We have no preference between these possibilities---we simply note
that the \cn\ sample could be another helpful data point in
distinguishing them.

@@

\subsection{How did the halo form?}
\label{subsec:origin}

@@

\subsection{Mass differences between center and outer reaches?}

@@

\subsection{Ways of doing this at different ages}
Some final discussion is warranted on the applicability of our
approach more generally.  This work starts with a kinematically
identified population, and then uses photometry (Gaia HR diagrams;
TESS rotation) and spectroscopy (GALAH and Gaia-ESO lithium) to
confirm youth in the stars.  Other approaches are also possible.

One can start with photometric criteria (e.g., Gaia HR diagram of
everything within some distance; Zari+18), or with spectroscopic
criteria (CITE Berger+18), or with combinations thereof (e.g., CITE
Zerjal 2019, 2020, Zhou's work).  
It might even be possible to start using chemical abundance signatures
(CITE, GALAHDR3?).

>>>>

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		%\subfloat{
			\includegraphics[width=1\textwidth]{f5a.pdf}
		%	\includegraphics[width=0.59\textwidth]{f5b.pdf}
		%}
	\end{center}
	\vspace{-0.7cm}
  \caption{ {\bf Lithium in the core and halo of NGC\,2516.}
  Equivalent width of the 6708\,\AA\ doublet is shown
  versus extinction-corrected color for all kinematic \cn\ members
  with Gaia-ESO ($N=452$) or GALAH ($N=107$) spectra available.
  The GALAH spectra comprise slightly over half of the halo stars,
  due to the non-targeted selection function of that survey.
  The gray points are Keck-HIRES measurements of stars with detected
  transiting planets from the Kepler mission
  \citep{berger_identifying_2018}.
  Points with ${\rm EW}\approx0\,$m\AA\ are non-detections.
  \label{fig:lithiumcorehalo}
  }
\end{figure*}
\begin{figure*}[t]
	\begin{center}
		\leavevmode
		%\subfloat{
		\includegraphics[width=1\textwidth]{f5b.pdf}
		%	\includegraphics[width=0.59\textwidth]{f5b.pdf}
		%}
	\end{center}
	\vspace{-0.7cm}
	\caption{ {\bf Lithium and rotation in NGC\,2516.}
		Points show all kinematic \cn\ members with TESS rotation
		periods in the Gaia-ESO or GALAH surveys, and are colored by
		equivalent width.
	   Non-detections appear with ${\rm EW}\approx0\,$m\AA.
	The K-dwarfs shows a similar lithium-rotation correlation
		to what has been previously observed,
		and is discussed in Section~\ref{subsec:lithium}.
		\label{fig:lithiumrot}
	}
\end{figure*}

@@

Figure~\ref{fig:lithiumcorehalo} shows the concatenation of the
Gaia-ESO and GALAH results, with the points colored by whether the
stars are in the core or the halo of the cluster.  The GALAH spectra
span a color range of $0<(Bp-Rp)_0<1.5$, due to the $V<14$ brightness
cutoff of the survey.  The overall depletion of stars redder than
$(Bp-Rp)_0\gtrsim1.4$, {\it i.e.}, later than K4.5V (0.71\,$M_\odot$),
is visible relative to the earlier K dwarfs.  For comparison, we have
also plotted the Li EWs measured by \citet{berger_identifying_2018}
for planet-hosting stars in the Kepler field (we interpolated from
their effective temperatures using the \citet{pecaut_mamajek_2013}
table\footnote{\url{http://www.pas.rochester.edu/~emamajek/EEM_dwarf_UBVIJHK_colors_Teff.txt},
version 2019.3.22.}.
The field star distribution peaks at late F-type stars.
The majority of kinematically selected stars in the core and the halo
show lithium equivalent widths substantially in excess of these field
stars.

We can go one step further, and match our sample of kinematically
selected stars with spectra against the TESS rotators.
The result is shown in Figure~\ref{fig:lithiumrot}
At fixed stellar mass
(and age), the rapid rotators tend to show elevated lithium equivalent
widths. This effect is mostly apparent in the K dwarfs.  Similar
trends in the Pleiades were noted nearly three decades ago by
\citet{soderblom_evolution_1993}.  More recent studies have found
comparable results in the Pleiades, the Psc-Eri stream, and M\,35
\citep{bouvier_pleiades_lirot_2018,arancibia_2020,jeffries_m35_li_2020}.

% The fact that we see the same trend across the core and halo of \cn\ 
% {\it a)} supports the interpretation that the halo is coeval with the
% core, and {\it b)} suggests that the lithium-rotation correlation is
% not explained by environmental effects such as the density of the cluster
% during its embedded phase, but is instead tied to physical processes
% acting on all low-mass pre-main-sequence stars.

% TODO: this paragraph more or less parrots the processes described by
% Bouvier+20. are there others? is there a better way to describe
% these?

An important point regarding the lithium-rotation trend is that it has
been shown to carry over beyond just equivalent widths, and to the
actual abundance of lithium in the photosphere, A(Li).  A priori, one
might worry that the trend is tautological: perhaps some feature the
stellar equatorial velocity, or the microturbulent velocity at the
stellar surface, are larger for faster rotators and this could affect
the formation conditions of the lithium doublet.
%FIXME improve these words
Soderblom+1993 calculated the relevant LTE curves of growth for the
lithium line, and argued that this could not be the case.
Since then, LindAsplund+2009 have calculated the relevant non-LTE
correction, and has shown that it affects the abundances by
$<0.5\,{\rm dex}$.
The order of magnitude of the rotation-abundance trend is 2-3\,dex.

%     * OK, then to convert to an ABUNDANCE, follow Bouvier+2018. In other words,
%       just interpolate against the LTE curve of growth tables from
%       Soderblom+1993.
% 
%     * Nb. the Non-LTE correction has been shown to be <0.5 dex (but
%       >0.1 dex, for sun-like stars with Li abundances above 2 dex,
%       >which is the regime we're interested in for Pleaides and NGC
%       >2516).
%       CITE: Lind, Asplund, + 2009
%       https://ui.adsabs.harvard.edu/abs/2009A%26A...503..541L/abstract

Processes both internal and external to the star have been suggested to explain
the lithium-rotation trend (CITE see the recent review by Bouvier+20).
One explanation based in the stellar interior could be that the
convective mixing efficiency is anticorrelated with the surface
rotation (e.g., CITE Siess + Livio 1997, Baraffe+2017).  Another
possibility could be that stronger magnetic fields in the star's
interior inhibit convection (READ e.g., Ventura+98, Chabrier+07,
Somers + Pinsonneault 2014).  An external process that might also be
important is the effect of star-disk magnetic locking during the PMS
phase (CITE: magnetic braking).  Longer disk lifetimes would lead to
the star's outer convective zone being locked for longer while the
radiative core contracts.  The resulting differential rotation and
rotational mixing could drive the lithium depletion (CITE: Bouvier 08,
Eggenberger+12).
We have no preference between these possibilities---we simply note
that the \cn\ sample could be another helpful data point in
distinguishing them.

@@

\subsection{How did the halo form?}
\label{subsec:origin}
% Should we be calling the halo a tail?

@@

% \subsection{Mass differences between center and outer reaches?}

@@

\subsection{Ways of doing this at different ages}
Some final discussion is warranted on the applicability of our
approach more generally.  Given a kinematically
identified population, this study uses photometry (Gaia HR diagrams;
TESS rotation) and spectroscopy (GALAH and Gaia-ESO lithium) to
confirm youth in the stars.  Other approaches are also possible.

One can use photometric criteria (e.g., Gaia HR diagram of
everything within some distance; Zari+18), or spectroscopic
criteria (CITE Berger+18), or with combinations thereof (e.g., CITE
Zerjal 2019, 2020, Zhou's work).  
It might even be possible to start using chemical abundance signatures
(CITE, GALAHDR3?).
